A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex
More work.

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\widetilde{#1}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

\begin{report}
Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$).
\end{report}
Symbols starting with an uppercase letter (e.g.,
$\const{S}$, $\const{SCons}$) are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (with $n \ge 0$)
and a return type $\tau$. This connection can be expressed %compactly
as $\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$\begin{report},
which collapses to $\const{f} : \tau$ if $n = 0$\end{report}.
For a term $t$, the notation $t : \tau$ indicates that it has type $\tau$.
Functions are invoked in the standard way, with $\const{f}(t_1,\ldots,t_n)$
applying the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 :\nobreak \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols are called constants and can appear without
parentheses in terms.
%
Finally, the notation $\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$
and $\bigwedge_{\,i}\, \varphi_i$ abbeviates a conjunction
$\varphi_1 \mathrel\land \cdots \mathrel\land \varphi_n$.

@@

The decision procedure for the ground theory of (co)datatypes $\thD$ determines the
$\thD$-satisfiability of sets of ground constraints over a fixed
signature~$\Sig$. It is presented as a calculus, whose rules can be applied in
any order until saturation.

The derivation rules operate on a set $\Ec$ of equalities and disequalities
between $\Sig$-terms, where $\Ec$ is interpreted as a conjunction.
We write $\tEc$ to denote the set of $\Sig$-terms occurring in $\Ec$.
%We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
%We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_j( \const{C}_j(t_1,\ldots,t_n) )$
%are simplified to $t_k$. All constraints added to $\Ec$ are also implicitly
%normalized in this manner.

To simplify the presentation, we make the following assumptions about $\Sig$.
First, all codatatypes in $\Codata$ are corecursive. This is not restrictive
because noncorecursive codatatypes can be considered as nonrecursive
datatypes.
Second, all types $\tau \in \Nondata$ have infinite cardinality.
This is not overly restrictive, because each interpreted type $\tau \in \Nondata$
having finite cardinality~$n$
can be replaced by an enumeration datatype with $n$~nullary constructors.
Moreover, since the constraints are ground, they cannot entail an upper bound
on the cardinality of any uninterpreted type $\tau \in \Nondata$; thus,
we may without loss of completeness treat all other types as infinite.

The calculus consists of three sets of rules (\figuresname~\ref{fig:cc-rules} to
\ref{fig:split-rule}) corresponding to three successive phases. The first
phase computes the bidirectional closure of $\Ec$. The second phase makes
inferences based on acyclicity (for datatypes) and uniqueness (for
codatatypes). The third phase performs case splitting on constructors for
various terms in $\tEc$.
%Following the conventions from~\cite{},
%The derivation rules are given in \relax{guarded assignment form},
%where
A rule can be applied to $\Ec$ if all the specified preconditions are met.
The conclusion of a rule either describes equalities to be added to $\Ec$
or is $\bot$.
% (in which case we call an application of it \emph{nonterminal})
% (in which case we call an application of it \emph{terminal}).
A rule may have multiple conclusions separated by $\parallel$,
denoting nondeterministic branching. The rules associated with a phase are
applied until saturation (i.e., until no rule can be applied to extend~$\Ec$
further) before the next phase can begin.

@@

\begin{example}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of \rn{Update} rule applications is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item \noindent\rlap{Initial state:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{  \}$
\item \noindent\rlap{Update w.r.t.\ $\ec{x}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) \}$
\item \noindent\rlap{Update w.r.t.\ $\ec{y}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}$
\item \noindent\rlap{Update w.r.t.\ $\ec{x}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\; \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) ) \vthinspace\}$
\end{enumerate}
%
This indicates that the values for $x$ and $y$ in models of $\Ec$
are of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots ))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
$\Box$
\end{example}

@@

\begin{example}
Let $\const{C}$ be a constructor of type $\tau_1 \times \tau_2 \rightarrow \tau_2$,
and let $\Ec$ be the set $\{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After phase~1, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
We then construct $\Val$ using the aforementioned phases, after which we obtain:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l}
\Val \ec{x} & = &
\mu {\Varec{x}}. & \const{C}( {\Varec{y}}, & \mu {\Varec{\const{C}(y,x)}}. & \const{C}( {\Varec{y}}, {\Varec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = &
\mu {\Varec{\const{C}(y,x)}}. & \const{C}( {\Varec{y}}, & \mu {\Varec{x}}. & \const{C}( {\Varec{y}}, {\Varec{\const{C}(y,x)}} ) )
\end{array}\]
Notice that $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are syntactically equivalent for the renaming
$\{ {\Varec{x}} \mapsto {\Varec{\const{C}(y,x)}}, {\Varec{\const{C}(y,x)}} \mapsto {\Varec{x}} \}$,
and thus our calculus may infer $x \teq \const{C}(y,x)$.
Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$
are equivalent for any value of ${\Varec{y}}$.
$\Box$
\end{example}

\begin{figure}[t]
\normalsize
\centering
\begin{tabular}{c}
\(
\inferrule{
  \tau \in \Codata
  \quad
  \left|\tau\right| = 1
  \quad
  t, u : \tau \in \tEc
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Singleton}
\\[5\jot]
\(
\inferrule{
  t : \tau \in \tEc
  \quad
  ( s( t ) \in \tEc, s \in \Sel^\tau )
  \text{ or }
  ( \tau \in \Data, \tau \text{ finite } )
}{
  \Ec := \Ec,\: t \teq \const{C}_1( s^1_1( t ), \ldots, s^{a_1}_1( t ) ) \ror \cdots \ror \Ec := \Ec, t \teq \const{C}_n( s^1_n( t ), \ldots, s^{a_n}_n( t ) )
}
\)
\rn{Split}
\end{tabular}
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

@@

Given a $\mu$-term $t_1$ of the form $\mu x. \const{C}( u_1, \ldots, u_n )$ with subterm $t_2$,
we say $t_2$ is a \emph{bisimilar subterm} of $t_1$ if and only if it is of the form
$\mu y. \const{C}( v_1, \ldots, v_n )$ and $t_1\interp{ u_i } \sim t_1\interp{ v_i }$ for $i = 1,\ldots, n$.
We say $t_1$ is \emph{normal} if and only if it does not contain a strict bisimilar subterm
and all of its strict subterms are also normal.
For instance, the $\mu$-term $t_1$ of the form $\mu x. \const{C}( \mu y. \const{C}( y ) )$ is not normal,
since $\mu y. \const{C}( y )$ is a bisimilar subterm of $t_1$.
In particular, note that their arguments have the same interpretation in $t_1$,
that is,
$t_1\interp{\mu y. \const{C}( y )}^\emptyset =$
$\mu y. \const{C}( t_1\interp{y}^{\{y\}} ) =$
$\mu y. \const{C}( y )$
is syntactically equivalent to
$t_1\interp{y}^\emptyset =$
$\mu y. \const{C}( t_1\interp{y}^{\{y\}} ) =$
$\mu y. \const{C}( y )$.
The $\mu$-term $t_1$ of the form $\mu x. \const{C}( \mu y. \const{C}( x ) )$ is also not normal,
since $\mu y. \const{C}( x )$ is a bisimilar subterm of $t_1$,
noting that
$t_1\interp{\mu y. \const{C}( x )}^\emptyset =$
$\mu y. \const{C}( t_1\interp{x}^{\{y\}} ) =$
$\mu y. \const{C}( t_1\interp{\mu x. \const{C}( \mu y. \const{C}( x ) ) }^{\{y\}} ) =$
$\mu y. \const{C}( \mu x. \const{C}( t_1\interp{\mu y. \const{C}( x )}^{\{x,y\}} ) ) =$
$\mu y. \const{C}( \mu x. \const{C}( y ) )$
is syntactically equivalent to $t_1\interp{x}^\emptyset = t_1$ for the renaming $\{ x \mapsto y, y \mapsto x \}$.

@@

\begin{lemma}
\label{lem:mu-norm-arg}
If $\vec u$ are normal, then if $t = \nf{(\mu x. \const{C}( \vec u ) )} = \mu x. \const{C}( \vec w )$,
then $\vec u \vsim t\interp{\vec w}$.
\end{lemma}
\begin{proof}
%Since $u_i$ is normal, $w_i$ is the result of replacing all subterms $v$ of $u_i$ such that $t\interp{v} \sim t$ with $x$.
\rem{TODO}
\end{proof}

@@

\begin{example}
Let $\tau$ be a type in $\Codata$ with three constructors $\const{A}$, $\const{B}$ and $\const{C}$ each having type $\tau \rightarrow \tau$.
Let $\Ec$ be the set consisting of equalities
$\{$
$y_1 \teq \const{A}( y_2 )$,
$y_2 \teq \const{B}( x )$,
$z_1 \teq \const{C}( z_2 )$,
$z_2 \teq \const{A}( z_1 )$,
$w_1 \teq \const{A}( x )$,
$w_2 \teq \const{C}( w_3 )$,
$x \not\teq y_2$,
$\}$.
After running the calculus to saturation on $\Ec$, the mapping $\Val$ contains:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{2.0em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\Val \ec{y_1} & = & \mu {\Varec{y_1}}. & \const{A}( & \mu {\Varec{y_2}}. & \const{B}( {\Varec{x}} ) ) &
\Val \ec{w_1} & = & \mu {\Varec{w_1}}. & \const{A}( {\Varec{x}} )      & \\
\Val \ec{y_2} & = & \mu {\Varec{y_2}}. & \multicolumn{2}{l}{\const{B}( {\Varec{x}} )}      & &
\Val \ec{w_2} & = & \mu {\Varec{w_2}}. & \const{C}( {\Varec{w_3}} ) \\
\Val \ec{z_1} & = & \mu {\Varec{z_1}}. & \const{C}( & \mu {\Varec{z_2}}. & \const{A}( {\Varec{z_1}} ) ) &
\Val \ec{w_3} & = & {\Varec{w_3}} \\
\Val \ec{z_2} & = & \mu {\Varec{z_2}}. & \const{A}( & \mu {\Varec{z_1}}. & \const{C}( {\Varec{z_2}} ) ) &
\Val \ec{x} & = & {\Varec{x}} \\
\end{array}\]

>>>>

\newcommand\xend{{\hfill\footnotesize$\blacksquare$}}

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\kern.45em \parallel \kern.45em}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\widetilde{#1}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

\begin{report}
Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$).
\end{report}
Symbols starting with an uppercase letter (e.g.,
$\const{S}$, $\const{SCons}$) are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (with $n \ge 0$)
and a return type $\tau$. This connection can be expressed %compactly
as $\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$\begin{report},
which collapses to $\const{f} : \tau$ if $n = 0$\end{report}.
For a term $t$, the notation $t : \tau$ indicates that it has type $\tau$.
Functions are invoked in the standard way, with $\const{f}(t_1,\ldots,t_n)$
applying the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 :\nobreak \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols are called constants and can appear without
parentheses in terms.
%
Finally, $\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$\begin{report}
and $\bigwedge_{\,i}\, \varphi_i$ abbeviates a conjunction
$\varphi_1 \mathrel\land \cdots \mathrel\land \varphi_n$\end{report}.

@@

The decision procedure for the ground theory of (co)datatypes $\thD$ determines the
$\thD$-satisfiability of sets of ground constraints over a fixed
signature~$\Sig$. It is formulated as a calculus, whose rules can be applied in
any order until saturation.
%
The rules operate on a set $\Ec$ of equalities and disequalities
between $\Sig$-terms, where $\Ec$ is interpreted as a conjunction.
We write $\tEc$ to denote the set of $\Sig$-terms occurring in $\Ec$.
%We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
%We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_j( \const{C}_j(t_1,\ldots,t_n) )$
%are simplified to $t_k$. All constraints added to $\Ec$ are also implicitly
%normalized in this manner.

To simplify the presentation, we make the following assumptions about $\Sig$.
First, all codatatypes in $\Codata$ are corecursive. This is not restrictive
because noncorecursive codatatypes can be considered as nonrecursive
datatypes.
Second, all types $\tau \in \Nondata$ have infinite cardinality.
This is not overly restrictive, because each interpreted type $\tau \in \Nondata$
having finite cardinality~$n$
can be replaced by an enumeration datatype with $n$~nullary constructors.
Moreover, since the constraints are ground, they cannot entail an upper bound
on the cardinality of any uninterpreted type $\tau \in \Nondata$; thus,
we may without loss of generality or completeness treat all types in $\Nondata$ as infinite.

The calculus consists of three sets of rules, given in \figuresname~\ref{fig:cc-rules} to
\ref{fig:split-rule}, corresponding to three successive phases. The first
phase computes the bidirectional closure of $\Ec$. The second phase makes
inferences based on acyclicity (for datatypes) and uniqueness (for
codatatypes). The third phase performs case splitting on constructors for
various terms in $\tEc$.
%Following the conventions from~\cite{},
%The derivation rules are given in \relax{guarded assignment form},
%where
A rule can be applied to $\Ec$ if all the specified preconditions are met.
The conclusion of a rule either describes equalities to be added to $\Ec$
or is $\bot$.
% (in which case we call an application of it \emph{nonterminal})
% (in which case we call an application of it \emph{terminal}).
A rule may have multiple conclusions separated by $\parallel$,
denoting nondeterministic branching. The rules associated with a phase are
applied until saturation (i.e., until no rule can be applied to extend~$\Ec$
further) before the next phase can begin.

@@

\begin{example}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of \rn{Update} rule applications is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item \noindent\rlap{Initial state:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{  \}$
\item \noindent\rlap{Update w.r.t.\ $\ec{x}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) \}$
\item \noindent\rlap{Update w.r.t.\ $\ec{y}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}$
\item \noindent\rlap{Update w.r.t.\ $\ec{x}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\; \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) ) \vthinspace\}$
\end{enumerate}
%
This indicates that the values for $x$ and $y$ in models of $\Ec$
are of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots ))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
\xend
\end{example}

@@

\begin{example}
Let $\const{C}$ be a constructor of type $\tau_1 \times \tau_2 \rightarrow \tau_2$,
and let $\Ec$ be the set $\{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After phase~1, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
We then construct $\Val$ using the aforementioned phases, after which we obtain:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l}
\Val \ec{x} & = &
\mu {\Varec{x}}. & \const{C}( {\Varec{y}}, & \mu {\Varec{\const{C}(y,x)}}. & \const{C}( {\Varec{y}}, {\Varec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = &
\mu {\Varec{\const{C}(y,x)}}. & \const{C}( {\Varec{y}}, & \mu {\Varec{x}}. & \const{C}( {\Varec{y}}, {\Varec{\const{C}(y,x)}} ) )
\end{array}\]
Notice that $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are syntactically equivalent for the renaming
$\{ {\Varec{x}} \mapsto {\Varec{\const{C}(y,x)}}, {\Varec{\const{C}(y,x)}} \mapsto {\Varec{x}} \}$,
and thus our calculus may infer $x \teq \const{C}(y,x)$.
Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$
are equivalent for any value of ${\Varec{y}}$.
\xend
\end{example}

\begin{figure}[t]
\normalsize
\centering
\begin{tabular}{c}
\(
\inferrule{
  \tau \in \Codata
  \quad
  \left|\tau\right| = 1
  \quad
  t, u : \tau \in \tEc
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Singleton}
\\[5\jot]
\(
\inferrule{
  t : \tau \in \tEc
  \quad
  ( s( t ) \in \tEc \text{ and } s \in \Sel^\tau )
  \text{ or }
  ( \tau \in \Data \text{ and } \tau \text{ finite} )
}{
  \Ec := \Ec,\: t \teq \const{C}_1(\const s^1_1( t ), \ldots,\const s^{n_1}_1( t ) ) \ror \cdots \ror \Ec := \Ec,\: t \teq \const{C}_m(\const s^1_m( t ), \ldots,\const s^{n_m}_m( t ) )
}
\)
\rn{Split}
\end{tabular}
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

@@

Given a $\mu$-term $t_1$ of the form $\mu x. \const{C}( u_1, \ldots, u_n )$ with subterm $t_2$,
we say $t_2$ is a \emph{bisimilar subterm} of $t_1$ if and only if it is of the form
$\mu y. \const{C}( v_1, \ldots, v_n )$ and $t_1\interp{ u_i } \aequiv t_1\interp{ v_i }$ for $i = 1,\ldots, n$.
We say $t_1$ is \emph{normal} if and only if it does not contain a strict bisimilar subterm
and all of its strict subterms are also normal.
For instance, the $\mu$-term $t_1$ of the form $\mu x. \const{C}( \mu y. \const{C}( y ) )$ is not normal,
since $\mu y. \const{C}( y )$ is a bisimilar subterm of $t_1$.
In particular, note that their arguments have the same interpretation in $t_1$,
that is,
$t_1\interp{\mu y. \const{C}( y )}^\emptyset =$
$\mu y. \const{C}( t_1\interp{y}^{\{y\}} ) =$
$\mu y. \const{C}( y )$
is syntactically equivalent to
$t_1\interp{y}^\emptyset =$
$\mu y. \const{C}( t_1\interp{y}^{\{y\}} ) =$
$\mu y. \const{C}( y )$.
The $\mu$-term $t_1$ of the form $\mu x. \const{C}( \mu y. \const{C}( x ) )$ is also not normal,
since $\mu y. \const{C}( x )$ is a bisimilar subterm of $t_1$,
noting that
$t_1\interp{\mu y. \const{C}( x )}^\emptyset =$
$\mu y. \const{C}( t_1\interp{x}^{\{y\}} ) =$
$\mu y. \const{C}( t_1\interp{\mu x. \const{C}( \mu y. \const{C}( x ) ) }^{\{y\}} ) =$
$\mu y. \const{C}( \mu x. \const{C}( t_1\interp{\mu y. \const{C}( x )}^{\{x,y\}} ) ) =$
$\mu y. \const{C}( \mu x. \const{C}( y ) )$
is syntactically equivalent to $t_1\interp{x}^\emptyset = t_1$ for the renaming $\{ x \mapsto y, y \mapsto x \}$.

@@

\begin{lemma}
\label{lem:mu-norm-arg}
If $\vec u$ are normal, then if $t = \nf{(\mu x. \const{C}( \vec u ) )} = \mu x. \const{C}( \vec w )$,
then $\vec u \vsim t\interp{\vec w}$.
\end{lemma}
\begin{proof}
%Since $u_i$ is normal, $w_i$ is the result of replacing all subterms $v$ of $u_i$ such that $t\interp{v} \aequiv t$ with $x$.
\rem{TODO}
\end{proof}

@@

\begin{example}
Let $\tau$ be a type in $\Codata$ with three constructors $\const{A}$, $\const{B}$ and $\const{C}$ each having type $\tau \rightarrow \tau$.
Let $\Ec$ be the set
$\{$
$y_1 \teq \const{A}( y_2 )$,
$y_2 \teq \const{B}( x )$,
$z_1 \teq \const{C}( z_2 )$,
$z_2 \teq \const{A}( z_1 )$,
$w_1 \teq \const{A}( x )$,
$w_2 \teq \const{C}( w_3 )$,
$x \not\teq y_2$,
$\}$.
After running the calculus to saturation on $\Ec$, the mapping $\Val$ contains:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{2.0em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\Val \ec{y_1} & = & \mu {\Varec{y_1}}. & \const{A}( & \mu {\Varec{y_2}}. & \const{B}( {\Varec{x}} ) ) &
\Val \ec{w_1} & = & \mu {\Varec{w_1}}. & \const{A}( {\Varec{x}} )      & \\
\Val \ec{y_2} & = & \mu {\Varec{y_2}}. & \multicolumn{2}{l}{\const{B}( {\Varec{x}} )}      & &
\Val \ec{w_2} & = & \mu {\Varec{w_2}}. & \const{C}( {\Varec{w_3}} ) \\
\Val \ec{z_1} & = & \mu {\Varec{z_1}}. & \const{C}( & \mu {\Varec{z_2}}. & \const{A}( {\Varec{z_1}} ) ) &
\Val \ec{w_3} & = & {\Varec{w_3}} \\
\Val \ec{z_2} & = & \mu {\Varec{z_2}}. & \const{A}( & \mu {\Varec{z_1}}. & \const{C}( {\Varec{z_2}} ) ) &
\Val \ec{x} & = & {\Varec{x}} \\
\end{array}\]

To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, ${\Varec{x}}$ and ${\Varec{w_3}}$.
The content of the set $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$ is $\alpha$-equivalent to:
\[%\begin{array}{c}
\{
\mu x. \const{A}( \const{B}( x )),
\const{A}( \mu y. \const{B}( y )),
\mu x. \const{B}( x ),
\mu x. \const{C}( \const{A}( x )),
\mu x. \const{A}( \const{C}( x )),
\mu x. \const{A}( x )
\}
%\end{array}
\]
Consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$,
for instance, $x$ is interpretted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for ${\Varec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\M$ to be inconsistent.
For instance, say we assign $\mu {\Varec{x}}. \const{B}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\mu {\Varec{y_2}}. \const{B}( \mu {\Varec{x}}. \const{B}( {\Varec{x}} ) )$,
which has normal form $\mu {\Varec{y_2}}. \const{B}( {\Varec{y_2}} )$,
which is $\alpha$-equivalent to $\mu {\Varec{x}}. \const{B}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.
Inversely, if the value assigned to ${\Varec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$,
then the values in the range of $\Val$ remain $\alpha$-disequivalent.
For instance, we may assign values such as 
$\mu {\Varec{x}}. \const{C}( {\Varec{x}} )$,
$\mu {\Varec{x}}. \const{C}( \mu x_1. \const{A}( x_1) )$, or
$\mu {\Varec{x}}. \const{B}( \const{A}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Legal substitutions for $v_{\ec{x}}$ may cause the range of $\Val$ to contain non-normal terms.
After applying the substitution for the third term mentioned above,
$\Val \ec{w_1}$ is $\mu {\Varec{w_1}}. \const{A}( \mu {\Varec{x}}. \const{B}( \const{A}( {\Varec{x}} ) ) )$,
which has normal form
$\mu {\Varec{w_1}}. \const{A}( \mu {\Varec{x}}. \const{B}( {\Varec{w_1}} ) )$.
\xend
\end{example}

%We first state the following properties of $\Val^\ast$.
%where we write $\muvar( t )$ to denote $t$ if $t$ is a variable, or $x$ if $t$ is of the form $\mu x. u$.
